---
title: "Report"
author: "Fabian Krügel"
date: "2024-05-24"
output:
  word_document: default
  html_document: default
---

# Capstone Project: Analyzing Global Health Data

## Data Extraction and Cleaning

```{r, echo=FALSE}
library(tidyverse)
library(httr)
library(rvest)
library(jsonlite)
library(countrycode)

# Get the list with all the variables that are available (indicators)
response <- httr::GET(
  url = 'https://ghoapi.azureedge.net/api/Indicator',
  verbose()
)
```


```{r, echo=FALSE}}
# Make a data frame of the list
cnt <- content(response, as = "parsed")
dat <- tibble(
  code = map_chr(cnt$value, 1), 
  name = map_chr(cnt$value, 2), 
  language = map_chr(cnt$value, 3)
)
```

```{r, echo=FALSE}}
# Download data from one indicator (Life expectancy): 
dat[dat$code == "WHOSIS_000001",]
response <- httr::GET(
  url = 'https://ghoapi.azureedge.net/api/WHOSIS_000001',
  verbose()
)
cnt_life_exp <- content(response, as = "parsed")
```

```{r, echo=FALSE}}
# Function to extract elements and convert if necessary
extract_chr <- function(x, index) {
  val <- x[[index]]
  if (is.null(val)) return(NA_character_)
  as.character(val)
}

extract_dbl <- function(x, index) {
  val <- x[[index]]
  if (is.null(val)) return(NA_real_)
  as.numeric(val)
}
```

```{r, echo=FALSE}}
# Get the list of indicators
response <- httr::GET(url = 'https://ghoapi.azureedge.net/api/WHOSIS_000001', verbose())

# Parse the response as text
raw_content <- content(response, as = "text")

# Convert the JSON text to a list
json_data <- fromJSON(raw_content, flatten = TRUE)

# Inspect the structure of the data
str(json_data)
```

```{r, echo=FALSE}}
# Print a subset of the data to understand its structure
print(json_data$value[1:5])  # Print the first 5 entries to inspect
```

```{r, echo=FALSE}}
# Create the tibble with the extracted data
dat_lifeexp <- tibble(
  id = map_chr(cnt_life_exp$value, ~ extract_chr(.x, "Id")),         # Extracts the first element (`Id`) from all the lists inside the cnt$value list
  country = map_chr(cnt_life_exp$value, ~ extract_chr(.x, "SpatialDim")),    # Extracts the 4th element (`SpatialDim`, country acronym)
  year = map_dbl(cnt_life_exp$value, ~ extract_dbl(.x, "TimeDim")),       # Extracts the year by name
  sex = map_chr(cnt_life_exp$value, ~ extract_chr(.x, "Dim1")),  # Assuming 'Dim1' is the
  life_expectancy = map_dbl(cnt_life_exp$value, ~ extract_dbl(.x, "NumericValue"))
)
```

```{r, echo=FALSE}}
# Filter for "Both sexes"
dat_lifeexp <- dat_lifeexp %>%
  filter(sex == "SEX_BTSX") %>%
  select(-sex)  
```

```{r, echo=FALSE}}
# Convert country codes to country names using countrycode package
dat_lifeexp <- dat_lifeexp %>%
  mutate(country_name = countrycode(country, "iso3c", "country.name"))
```

```{r, echo=FALSE}}
# View the updated tibble
print(dat_lifeexp)
```

```{r, echo=FALSE}}
# Remove rows where country_name is NA
dat_lifeexp <- dat_lifeexp %>%
  filter(!is.na(country_name))
dat_lifeexp

# Save the data to a CSV file
#write_csv(dat_lifeexp, "data/life_expectancy.csv")
```



## Shiny App

```{r}
# Create the Shiny app
library(shiny)


# Define UI for the application
ui <- fluidPage(
  titlePanel("Life Expectancy Over Time"),
  sidebarLayout(
    sidebarPanel(
      selectInput("country", "Select Country/Countries:", choices = sort(unique(dat_lifeexp$country_name)), multiple = TRUE)
    ),
    mainPanel(
      plotOutput("lifeExpPlot")
    )
  )
)

# Define server logic required to draw the plot
server <- function(input, output) {
  output$lifeExpPlot <- renderPlot({
    filtered_data <- dat_lifeexp %>% filter(country_name %in% input$country)
    ggplot(filtered_data, aes(x = year, y = life_expectancy, color = country_name)) + 
      geom_line(size = 1) +
      geom_point(size = 2) +
      labs(title = "Life Expectancy Over Time",
           x = "Year",
           y = "Life Expectancy",
           color = "Country") +
      theme_minimal()
  })
}

# Run the application 
#shinyApp(ui = ui, server = server)
```


## Plots

```{r}
# Plotting life expectancy over time:
ggplot(dat_lifeexp, aes(x = year, y = life_expectancy)) + 
  geom_smooth(aes(group = country), size = 0.1, se = FALSE, color = "black") + 
  geom_smooth(size = 2) +
  labs(title = "Life Expectancy Over Time", x = "Year", y = "Life Expectancy")


# Life Expectancy Distribution in 2019
life_exp_2019 <- dat_lifeexp %>% filter(year == 2019)
median_life_exp <- median(life_exp_2019$life_expectancy, na.rm = TRUE)

ggplot(life_exp_2019, aes(x = life_expectancy)) +
  geom_histogram(binwidth = 1, fill = "blue", alpha = 0.7) +
  geom_vline(aes(xintercept = median_life_exp), color = "red", linetype = "dashed", size = 1) +
  annotate("text", x = median_life_exp + 2, y = max(table(life_exp_2019$life_expectancy)), 
           label = paste("Median:", round(median_life_exp, 2)), color = "red") +
  labs(title = "Distribution of Life Expectancy in 2019", x = "Life Expectancy", y = "Frequency") +
  theme_minimal()


# Top and Bottom 5 Countries by Life Expectancy in 2015
top5 <- dat_lifeexp %>% filter(year == 2019) %>% top_n(5, life_expectancy)
bottom5 <- dat_lifeexp %>% filter(year == 2019) %>% top_n(-5, life_expectancy)
top_bottom <- bind_rows(top5, bottom5)

ggplot(top_bottom, aes(x = reorder(country_name, life_expectancy), y = life_expectancy, fill = country_name)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top and Bottom 5 Countries by Life Expectancy in 2015", x = "Country", y = "Life Expectancy") +
  theme_minimal() +
  theme(legend.position = "none")


# Change in Life Expectancy Over Time for Selected Countries
selected_countries <- c("Italy", "Japan", "United States", "India", "Brazil")

ggplot(dat_lifeexp %>% filter(country_name %in% selected_countries), aes(x = year, y = life_expectancy, color = country_name)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Change in Life Expectancy Over Time", x = "Year", y = "Life Expectancy", color = "Country") +
  theme_minimal()
```


## Conclusion

In this capstone project, I analyzed global health data, specifically focusing on life expectancy across various countries. Using data from the World Health Organization (WHO) performed the following key steps:

1. **Data Extraction and Cleaning:** 
   - I extracted life expectancy data using WHO's API (gho-odata-api), ensuring that the data was correctly parsed and cleaned.
   - I filtered the data to include only the life expectancy for both sexes, providing a comprehensive overview.

2. **Shiny App Development:**
   - I developed a simple Shiny application that allows users to interactively visualize the life expectancy trends over time for all countries.
   - The app provides an intuitive interface for selecting countries and viewing their respective life expectancy data.

3. **Data Visualization:**
   - I created various plots to illustrate the life expectancy trends and distributions.
   - These visualizations include:
     - A smooth trend of life expectancy over time for all countries.
     - A histogram depicting the distribution of life expectancy in 2019, with the median highlighted.
     - A bar chart showing the top and bottom 5 countries by life expectancy in 2019.
     - A line plot highlighting the changes in life expectancy over time for selected countries.

### Key Findings:

- The visualizations reveal significant disparities in life expectancy across different countries.
- There has been a general upward trend in life expectancy over the years, although the rate of increase varies by country and cannot be observed in each country.
- The top and bottom 5 countries by life expectancy in 2019 highlight the extremes, offering a perspective on the best and worst health outcomes.

### Future Work:

To build upon this project, the following steps are recommended:

- **Incorporate Additional Indicators:** Expand the analysis to include other health indicators such as mortality rates, disease prevalence, and healthcare access.
- **Regional Analysis:** Conduct a more detailed regional analysis to understand the underlying factors contributing to differences in life expectancy.
- **Predictive Modeling:** Develop predictive models to forecast future life expectancy trends based on historical data.
compare the life expectancy data with political (e.g. democracy index) and economic indicators (GDP per capita) to identify potential correlations.

Overall, this project demonstrates the power of data analytics and visualization in understanding global health trends. By leveraging interactive tools like Shiny apps, we can make data more accessible and insightful.


### Remark
My original plan was to perform a more complex sentiment analysis on the development of the sentiment of the term "single payer health insurance" over time. I wanted to scrape and analyse the text data of the Guardian, from other news pages, Google Scholar, and Pubmed. However, I had difficulties to receive usable data. Therefore, I decided to perform this more more simple analysis on life expectancy data. Unfortunately, I was not able to finish properly finish this project since for example knitting the document did not work properly. I am sorry for that. I addition I would have liked to include  more variables in the analysis to make the shiny app more interesting. 


